---
title: "README"
author: "Taiki Sakai"
date: "September 26, 2015"
output: html_document
---

### Script Explanation

The first step in our script is to extract the names of the variables from the
features document, and then transform these names so that they fit with R standard
naming conventions and so that ambiguous parts of the name are more explanatory.

```{r eval=FALSE}
features <- read.table("UCI HAR Dataset/features.txt")
fixedFeat <- gsub("\\()","",features[,2])
fixedFeat <- gsub("-","_",fixedFeat)
fixedFeat <- gsub("^f","freq",fixedFeat)
fixedFeat <- gsub("^t","time",fixedFeat)
fixedFeat <- gsub("Acc","Accel",fixedFeat)
fixedFeat <- gsub("Mag","Magnitude",fixedFeat)
```

fixedFeat stores a new versions of the names where we have removed the () and changed -
to _ to be in line with R column naming conventions.

We then make the names more descriptive by replacing t and f with time and freq, and expanding
Acc and Mag to Accel and Magnitude.

Next we read in our datasets. The X tables contain our data, so we will name the columns using the 
fixedFeat vector we just created.

```{r eval=FALSE}
Xtest <- read.table("UCI HAR Dataset/test/X_test.txt", col.names=fixedFeat)
Xtrain <- read.table("UCI HAR Dataset/train/X_train.txt", col.names=fixedFeat)
Ytest <- read.table("UCI HAR Dataset/test/y_test.txt")
Ytrain <- read.table("UCI HAR Dataset/train/y_train.txt")
testSub <- read.table("UCI HAR Dataset/test/subject_test.txt")
trainSub <- read.table("UCI HAR Dataset/train/subject_train.txt")
actLabels <- read.table("UCI HAR Dataset/activity_labels.txt")
```

The Y tables contain the activity type indicators as a 1-6 for each observation. The subject tables contain
the subject ID for each observation. The activity labels table tells us how to convert the 1-6 to an activity name.

Before we merge the test and training datasets, we will filter the existing data so that our merged object is smaller.
We begin by creating a logical vector to find column names containing "mean" and "std" so that we only have data
for the means and standard deviations. We subtract off the meanFreq values that get caught by this grep as they do not
have a corresponding std value and thus do not seem to be of the same sort as the other mean values. We also are not 
taking the values at the end of the dataset of the form angle(gravityMean ,... ) as these are angle measurements. 
These are missed by the grep because it is case sensitive.

```{r, eval=FALSE}
meanstd <- as.logical(grepl("mean", fixedFeat) + 
                            grepl("std", fixedFeat) - 
                            grepl("meanFreq", fixedFeat))
```

Next simply apply this logical vector to the datasets so that we have smaller tables containing only mean and std data.

```{r, eval=FALSE}
Xtest <- Xtest[,meanstd]
Xtrain <- Xtrain[,meanstd]
```

Now we create new columns on our test and training sets for the SubjectID and Activity variables. We must
do this before merging so that the correct IDs are given to the correct observations. The activity names
are created by casting the 1-6 values as a factor variable with the corresponding labels read in from the
activity text file. 

```{r, eval=FALSE}
Xtest$SubjectID <- testSub$V1
Xtrain$SubjectID <- trainSub$V1

Xtest$ActivityName <- factor(Ytest$V1,labels = actLabels$V2)
Xtrain$ActivityName <- factor(Ytrain$V1, labels=actLabels$V2)
```

Now we load the Data Table package for easier calculations, and we merge the two datasets together.

```{r, eval=FALSE}
library(data.table)
merged <- data.table(rbind(Xtrain, Xtest))
```

Now we create our tidy data table of the means of each of the columns for each SubjectID, ActivityName pairs.
This is where Data Tables make things much simpler. The .SD gives a subset of the datatable for each group that
we group by (see documentation for a more in depth explanation of .SD) so the lapply command allows us to easily 
calculate the mean for every column.

```{r, eval=FALSE}
tidyMeans<- merged[, lapply(.SD,mean), by=c("SubjectID","ActivityName")]
```

All thats left is to order by SubjectID to make a more logical ordering of our data and output to a text file.

```{r, eval=FALSE}
tidyMeans<- tidyMeans[order(SubjectID),]

write.table(tidyMeans, "TidyData.txt", row.names=FALSE)
```